Ecosystem size and complexity dictate riverine biodiversity

### 2021-01-25 manuscript.Rmd
Update manuscript.Rmd

First, we theoretically analyzed the influences of ecosystem size and branching complexity on $\gamma$ diversity. We depicted branching ecosystems as a spatial network of connected habitat patches or local communities[@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014a] (**Figure 1**) where two factors determine mean environmental conditions at each habitat patch: (1) headwater environments (the most upstream habitat patch) and (2) local environmental noise. Environmental values at headwaters are drawn randomly from a normal distribution and propagate downstream with local environmental noise (i.e., a spatial autoregressive process with white noise). These environmental values recurrently 'mix' at confluences considering the relative size of joining tributaries (see Methods). Therefore, our network-generation procedure resembles natural processes of how branching river networks create diverse habitats in a metacommunity. We then simulated metacommunity dynamics in the theoretical branching networks using a general metacommunity framework[@thompsonProcessbasedMetacommunityFramework2020]. In this model, 50 species with different abiotic niches (optimum and width) disperse along a network and compete for resources with varied strengths in temporally dynamic environments (see Methods for model details).

Given the results of extensive sensitivity analysis (**Supplementaly Information**), we considered `r n_scenario` simulation scenarios, which comprised a combination of four landscape and eight ecological scenarios. We distinguished landscape scenarios by setting four combinations of environmental variation at headwaters ($\sigma_h =$ `r unique(dat_sim$sd_env_source)`) and the degree of local environmental noise ($\sigma_l =$ `r unique(dat_sim$sd_env_lon)`), both of which are defined as standard deviations of normal distributions. When $\sigma_h > \sigma_l$, branching produces greater habitat heterogeneity because headwaters are the primary source of environmental variation. This landscape scenario should reproduce natural patterns of habitat heterogeneity, in which environmental conditions differ greatly among tributaries [@chezikRiverNetworksDampen2017; @mooreEmergentStabilityLarge2015; @petersonModellingDendriticEcological2013]. In the meantime, when $\sigma_h \le \sigma_l$, local environmental noise masks environmental variation among tributaries, leading to minimal influences of branching on habitat heterogeneity. This scenario may reflect human-modified landscapes where the physical or biological distinctiveness of tributaries has been compromised due to human activities[@mooreSynchronizationPortfolioPerformance2010; @poffHomogenizationRegionalRiver2007; @teruiMetapopulationStabilityBranching2018]. Thus, the inequality between $\sigma_h$ and $\sigma_l$ creates contrasting patterns of habitat heterogeneity within a network. We also considered eight ecological scenarios that differ in dispersal distance (controlled by the rate parameter of an exponential dispersal kernel), dispersal probability, and the maximum value of interspecific competition strength (see Methods). These simulation scenarios were capable of reproducing common spatial patterns of local species richness in rivers, corroborating the appropriateness of our choice in parameter combinations (Supplementary Information). Under each simulation scenario, we simulated 1000 time steps of metacommunity dynamics in `r max(dat_sim$n_rep)` branching networks with the gradients of ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`).

@@

The proposed theory provided important insights into how ecological communities are structured in branching networks; however, empirically testing the predictions is extremely difficult because it requires metacommunity-level replications. To confront this logistical challenge, we compiled fish community data that span across two geographic regions: the Hokkaido island in Japan and the midwestern US (Midwest). These regions are located in comparable ranges of latitude but support distinct fish communities (**Figure 1**). Therefore, this data set provides an excellent opportunity to examine the generality of our theoretical predictions. After careful data selection (see Methods and Supplementary Information), we estimated $\alpha$, $\beta$, and $\gamma$ diversity (asymptotic species richness; Methods) at `r nrow(wsd_subset[[1]]) + nrow(wsd_subset[[2]])` watersheds (`r nrow(wsd_subset[[1]])` in Hokkaido and `r nrow(wsd_subset[[2]])` in the Midwest), each of which comprised $\ge$ 10 sites of presence-absence fish community data (a total of `r nrow(point_subset[[1]]) + nrow(point_subset[[2]])` sites). We considered watersheds as independent replicates of metacommunities if the river network flows into one of the following: the ocean, a large lake/reservoir ($\ge$ 10 km² in the areal area), or a large river that may represent lentic habitats ($\ge$ 5000 km² in the watershed area). We combined this data set with geospatial information (watershed area, branching probability, annual mean temperature, annual cumulative precipitation, and % forest) to examine potential influences of macro-scale drivers of species diversity. Using this data set, we developed global and region-specific models for each of the diversity measures ($\alpha$, $\beta$, and $\gamma$ diversity) to examine whether observed patterns are consistent across the two geographic regions. In the global model, we assumed that effects of ecosystem size (watershed area) and complexity (branching probability) are constant across the two regions (i.e., fixed slopes). Meanwhile, the region-specific model assumes region-specific slopes of ecosystem size and complexity by including interaction terms with the dummy binary variable of the region (Hokkaido = 0; Midwest = 1). We compared the performance of these competing models using the Bayes factor, a measure of the strength of evidence in favor of one model over the alternative.

Despite the substantial difference in fish fauna between the study regions, the estimated Bayes factor in favor of the global model was `r round(fit$gamma$bayes_factor, 1)` for $\gamma$ diversity; this result strongly supports the consistent effects of ecosystem size and complexity on $\gamma$ diversity across the two biogeographic regions. We found patterns consistent with our theoretical predictions (**Table 1 and Figure 4**). The estimated $\gamma$ diversity increased with increasing watershed area (ecosystem size) and branching probability (ecosystem complexity) across watersheds. These effects remained significant even after controlling for the potential influences of other environmental factors (precipitation, temperature, and land use) (**Table 1**). Overall, our statistical analysis provides strong empirical evidence that ecosystem size and complexity jointly scale riverine biodiversity at the metacommunity level.

Similarly, we found weak to moderate supports for the global models of $\alpha$ and $\beta$ diversity (Bayes factor: `r round(fit$alpha$bayes_factor, 1)` and `r round(fit$beta$bayes_factor, 1)`, respectively). In both regions, $\beta$ diversity responded significantly to ecosystem size and complexity while $\alpha$ diversity showing a vague response to these ecosystem properties (**Table 1 and Figure 4**). In our simulations, this pattern has emerged under the scenarios with dispersal limitation, which elegantly matches the previous observations of stream fish movement. Direct (mark-recapture) and indirect observations (e.g., genotyping) recurrently revealed the restricted movement of stream fish, typically limited to several tens to hundreds of meters in distance[@comteFishDispersalFlowing2018; @radingerPatternsPredictorsFish2014; @rodriguezRestrictedMovementStream2002]. The reciprocal agreement of theoretical and empirical patterns provides indirect but convincing evidence that dispersal limitation plays a key role in driving the associations between $\gamma$ diversity and ecosystem properties in rivers.

@@

where $r_{ix}(t)$ is the reproductive rate of species $i$ given the environmental condition at patch $x$ and time $t$, $r_{0,i}$ the maximum reproductive rate of species $i$, $K_{x}$ the carrying capacity at patch $x$, $\alpha_{ij}$ the competition coefficient between species $i$ and $j$, and $S$ the number of species in a metacommunity. $K_x$ was expressed as a function of the number of upstream contributing patches $A_x$ ($K_x = 100A_x$) to represent increasing habitat size with increasing upstream watershed area[@altermattRiverNetworkProperties2013; @teruiThreeEcologicalFactors2016]. The parameter $\alpha_{ij}$ is expressed as the strength of interspecific competition relative to that of intraspecific competition such that interspecific competition is greater than intraspecific competition if $\alpha_{ij}$ > 1 (intraspecific competition coefficient was set constant as $\alpha_{ii} = 1$). $\alpha_{ij}$ was drawn randomly from a uniform distribution as $\alpha_{ij} \sim Unif(0, \alpha_{max})$. The density-independent reproductive rate $r_{ix}(t)$ is affected by abiotic environments (non-consumable) and determined by the following Gaussian function:

@@

We compiled fish data at two biogeographic regions: the Hokkaido island, Japan and the Midwest, US. For Hokkaido, we used data from the Hokkaido Freshwater Fish Database HFish[@fukushimaHokkaidoFreshwaterFish2011; @fukushimaModellingEffectsDams2007], monitoring data at protected watersheds[@comteRivFishTIMEGlobalDatabase2021; @teruiMetapopulationStabilityBranching2018], and primary data collected by the authors[@miyazakiTemporalDynamicsFluvial2016; @teruiThreeEcologicalFactors2016], which collectively cover the entire Hokkaido island. For the Midwest, we assembled fish community data collected by Iowa Department of Natural Resources, Illinois Environmental Protection Agency and Illinois Department of Natural Resources, Minnesota Pollution Control Agency, and Wisconsin Department of Natural Resources. We detailed procedures for initial data selection in Supplementary Information.

We considered watersheds as independent replicates of metacommunities if the river network flows into one of the following: the ocean, a large lake/reservoir ($\ge$ 10 km² in the areal area) or a large river that may represent lentic habitats ($\ge$ 5000 km² in the watershed area). Metacommunities with $\ge$ 10 sampling sites were used in our analysis because a small sample size inflates statistical uncertainties in estimating diversity metrics (Supplementary Information).

We estimated $\alpha$, $\beta$, and $\gamma$ diversity at each watershed. $\gamma$ diversity is the total species richness at each watershed. Since it is impossible to sample all species present in each watershed, we used asymptotic species richness as a proxy for $\gamma$ diversity. Asymptotic species richness can be interpreted as the estimated true species richness in a given area when estimated using spatial replicates of local community data within a metacommunity[@chaoRarefactionExtrapolationHill2014; @chaseEcosystemDecayExacerbates2020]. We estimated asymptotic species richness using sampling sites located in the same watershed (i.e., spatial replicates within a metacommunity) via the R package iNEXT[@hsiehINEXTPackageRarefaction2016]. We assessed the completeness of the observed species richness based on sample coverage, which is the proportion of observed species richness to asymptotic species richness. We excluded watersheds whose sample coverage was below 0.9 because the estimates of asymptotic species richness may contain non-negligible statistical uncertainty. $\alpha$ diversity was estimated by taking an average of local species richness at each site. $\beta$ diversity was estimated by dividing the estimated $\gamma$ diversity by $\alpha$ diversity.

@@

We collected data for watershed characteristics (watershed area and branching probability), climatic variables (annual mean air temperature and cumulative precipitation), and land use (% forest). We delineated watershed polygons and river lines (defined as $\ge$ 1 km^2^) using Hydrologically Adjusted Elevations of MERIT Hydro[@yamazakiMERITHydroHighResolution2019] with ArcMap 10.7 and SAGA 6.3.0 combined with the R package 'RSAGA'[@brenningRSAGASAGAGeoprocessing2018]. Watershed area was estimated based on the delineated watershed polygons using functions in the R package 'sf'[@pebesmaSimpleFeaturesStandardized2018]. Branching probability was estimated by fitting exponential distributions to histograms of branch length (unit: kilometer, river segment between successive confluences or a confluence and the outlet/upstream terminal) as $branch~length \sim Exp(\lambda)$. A derived statistical quantity $1 − e^{−\lambda}$ corresponds to the theoretical branching probability $P_b$. We estimated the spatial averages of air temperature and cumulative precipitation at a 1-km resolution for each watershed, which were extracted from WorldClim 1.4[@hijmansVeryHighResolution2005] via the R package 'raster.' We calculated % forest for each watershed using land use data in 2015 from Copernicus Global Land Service (100-m resolution)[@marcelbuchhornCopernicusGlobalLand2020].

@@

We developed linear models to examine the effects of macro-scale drivers on diversity metrics. Log-transformed diversity metrics ($\alpha$, $\beta$ and $\gamma$ diversity) at watershed $i$, $log_{10}~y_i$, are assumed to follow a normal distribution as $log_{10}~y_i \sim Normal(\mu_i, \sigma^2)$. The parameter $\mu_i$ was related to linear predictors, including log-transformed watershed area (ecosystem size; $log_{10}~WA$), log-branching probability (ecosystem complexity; $log_{10}~P_b$), region (coded as Hokkaido = 0, Midwest = 1; $region$), air temperature, precipitation, and percent of forest land use (logit-transformed). Although our primary focus of this analysis is to assess the effects of watershed area and branching probability, we included climate and land use variables to control these effects statistically.

Climate and land use differed clearly between the regions. Therefore, we used residuals of the simple linear regressions between the climate or land use variable (response variable) and region (explanatory variable) to avoid correlation between explanatory variables. The residuals represent deviations from the regional averages; positive values indicate values greater than the regional average while negative values indicating values smaller than the regional average.

@@

$$
\mu_i = \xi_0 + \xi_1~log_{10}~WA_i + \xi_2~log_{10}~P_{b,i} + \xi_3~region_i + \sum_k \xi_k x_{k,i}
$$
where $\xi_k$ is the intercept and regression coefficients for linear predictors. Meanwhile, the region-specific model assumes the effects of watershed area and branching probability differ between regions (i.e., interactions between region and watershed area/branching probability).

@@

# Acknowledgements

@@

>>>>

First, we theoretically analyzed the influences of ecosystem size and branching complexity on $\gamma$ diversity. We depicted branching ecosystems as a spatial network of connected habitat patches or local communities[@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014a] (**Figure 1**) where two factors determine mean environmental conditions at each habitat patch: (1) headwater environments (the most upstream habitat patch) and (2) local environmental noise. Environmental values at headwaters are drawn randomly from a normal distribution and propagate downstream with local environmental noise (i.e., a spatial autoregressive process with white noise). These environmental values recurrently 'mix' at confluences considering the relative size of joining tributaries (see **Methods**). Therefore, our network-generation procedure resembles natural processes of how branching river networks create diverse habitats in a metacommunity. We then simulated metacommunity dynamics in the theoretical branching networks using a general metacommunity framework[@thompsonProcessbasedMetacommunityFramework2020]. In this model, 50 species with different abiotic niches (optimum and width) disperse along a network and compete for resources with varied strengths in temporally dynamic environments (see **Methods** for model details).

Given the results of extensive sensitivity analysis (**Supplementaly Information**), we considered `r n_scenario` simulation scenarios, which comprised a combination of four landscape and eight ecological scenarios. We distinguished landscape scenarios by setting four combinations of environmental variation at headwaters ($\sigma_h =$ `r unique(dat_sim$sd_env_source)`) and the degree of local environmental noise ($\sigma_l =$ `r unique(dat_sim$sd_env_lon)`), both of which are defined as standard deviations of normal distributions. When $\sigma_h > \sigma_l$, branching produces greater habitat heterogeneity because headwaters are the primary source of environmental variation. This landscape scenario should reproduce natural patterns of habitat heterogeneity, in which environmental conditions differ greatly among tributaries [@chezikRiverNetworksDampen2017; @mooreEmergentStabilityLarge2015; @petersonModellingDendriticEcological2013]. In the meantime, when $\sigma_h \le \sigma_l$, local environmental noise masks environmental variation among tributaries, leading to minimal influences of branching on habitat heterogeneity. This scenario may reflect human-modified landscapes where the physical or biological distinctiveness of tributaries has been compromised due to human activities[@mooreSynchronizationPortfolioPerformance2010; @poffHomogenizationRegionalRiver2007; @teruiMetapopulationStabilityBranching2018]. Thus, the inequality between $\sigma_h$ and $\sigma_l$ creates contrasting patterns of habitat heterogeneity within a network. We also considered eight ecological scenarios that differ in dispersal distance (controlled by the rate parameter of an exponential dispersal kernel), dispersal probability, and the maximum value of interspecific competition strength (see **Methods**). These simulation scenarios were capable of reproducing common spatial patterns of local species richness in rivers, corroborating the appropriateness of our choice in parameter combinations (**Supplementary Information**). Under each simulation scenario, we simulated 1000 time steps of metacommunity dynamics in `r max(dat_sim$n_rep)` branching networks with the gradients of ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`).

@@

The proposed theory provided important insights into how ecological communities are structured in branching networks; however, empirically testing the predictions is extremely difficult because it requires metacommunity-level replications. To confront this logistical challenge, we compiled fish community data that span across two geographic regions: the Hokkaido island in Japan and the midwestern US (Midwest). These regions are located in comparable ranges of latitude but support distinct fish communities (**Figure 1**). Therefore, this data set provides an excellent opportunity to examine the generality of our theoretical predictions. After careful data selection (see **Methods** and **Supplementary Information**), we estimated $\alpha$, $\beta$, and $\gamma$ diversity (asymptotic species richness; **Methods**) at `r nrow(wsd_subset[[1]]) + nrow(wsd_subset[[2]])` watersheds (`r nrow(wsd_subset[[1]])` in Hokkaido and `r nrow(wsd_subset[[2]])` in the Midwest), each of which comprised $\ge$ 10 sites of presence-absence fish community data (a total of `r nrow(point_subset[[1]]) + nrow(point_subset[[2]])` sites). We considered watersheds as independent replicates of metacommunities if the river network flows into one of the following: the ocean, a large lake/reservoir ($\ge$ 10 km² in the areal area), or a large river that may represent lentic habitats ($\ge$ 5000 km² in the watershed area). We combined this data set with geospatial information (watershed area, branching probability, annual mean temperature, annual cumulative precipitation, and fraction of forest) to examine potential influences of macro-scale drivers of species diversity. Using this data set, we developed global and region-specific models for each of the diversity measures ($\alpha$, $\beta$, and $\gamma$ diversity) to examine whether observed patterns are consistent across the two geographic regions. In the global model, we assumed that effects of ecosystem size (watershed area) and complexity (branching probability) are constant across the two regions (i.e., fixed slopes). Meanwhile, the region-specific model assumes region-specific slopes of ecosystem size and complexity by including interaction terms with the dummy binary variable of the region (Hokkaido = 0; Midwest = 1). We compared the support of these competing models using the Bayes factor, a measure of the strength of evidence in favor of one model over the alternative. In our definition (see **Methods**), the Bayes factor of $> 1$ indicates the support for the global model over the region-specific model.

Despite the substantial difference in fish fauna between the study regions, we found strong support for the global model in explaining $\gamma$ diversity (Bayes factor = `r round(fit$gamma$bayes_factor, 1)`): this result suggests the consistent effects of ecosystem size and complexity on $\gamma$ diversity across the two biogeographic regions. Furthermore, the observed patterns were in close agreement with our theoretical predictions (**Table 1 and Figure 4**). The estimated $\gamma$ diversity increased with increasing watershed area (ecosystem size) and branching probability (ecosystem complexity) across watersheds. These effects remained significant even after controlling for the potential influences of other environmental factors (precipitation, temperature, and land use) (**Table 1**). Overall, our statistical analysis provides strong empirical evidence that ecosystem size and complexity jointly scale riverine biodiversity at the metacommunity level.

Similarly, we found weak to moderate supports for the global models of $\alpha$ and $\beta$ diversity (Bayes factor = `r round(fit$alpha$bayes_factor, 1)` and `r round(fit$beta$bayes_factor, 1)`, respectively). In both regions, $\beta$ diversity responded significantly to ecosystem size and complexity while $\alpha$ diversity showing a weaker or a vague response to these ecosystem properties (**Table 1 and Figure 4**). In our simulations, this pattern has emerged under the scenarios with dispersal limitation, which elegantly matches the previous observations of stream fish movement. Direct (mark-recapture) and indirect observations (e.g., genotyping) recurrently revealed the restricted movement of stream fish, typically limited to several tens to hundreds of meters in distance[@comteFishDispersalFlowing2018; @radingerPatternsPredictorsFish2014; @rodriguezRestrictedMovementStream2002]. The reciprocal agreement of theoretical and empirical patterns provides indirect but convincing evidence that dispersal limitation plays a key role in driving the associations between $\gamma$ diversity and ecosystem properties in rivers.

@@

where $r_{ix}(t)$ is the reproductive rate of species $i$ given the environmental condition at patch $x$ and time $t$, $r_{0,i}$ the maximum reproductive rate of species $i$, $K_{x}$ the carrying capacity at patch $x$, $\alpha_{ij}$ the competition coefficient between species $i$ and $j$, and $S$ the number of species in a metacommunity. $K_x$ was expressed as a function of the number of upstream habitat patches $A_x$ ($K_x = 100A_x$) to represent increasing habitat size with increasing upstream watershed area[@altermattRiverNetworkProperties2013; @teruiThreeEcologicalFactors2016]. The parameter $\alpha_{ij}$ is expressed as the strength of interspecific competition relative to that of intraspecific competition such that interspecific competition is greater than intraspecific competition if $\alpha_{ij}$ > 1 (intraspecific competition coefficient was set constant as $\alpha_{ii} = 1$). $\alpha_{ij}$ was drawn randomly from a uniform distribution as $\alpha_{ij} \sim Unif(0, \alpha_{max})$. The density-independent reproductive rate $r_{ix}(t)$ is affected by abiotic environments (non-consumable) and determined by the following Gaussian function:

@@

We compiled fish data at two biogeographic regions: the Hokkaido island, Japan and the Midwest, US. For Hokkaido, we used data from the Hokkaido Freshwater Fish Database HFish[@fukushimaHokkaidoFreshwaterFish2011; @fukushimaModellingEffectsDams2007], monitoring data at protected watersheds[@comteRivFishTIMEGlobalDatabase2021; @teruiMetapopulationStabilityBranching2018], and primary data collected by the authors[@miyazakiTemporalDynamicsFluvial2016; @teruiThreeEcologicalFactors2016], which collectively cover the entire Hokkaido island. For the Midwest, we assembled fish community data collected by Iowa Department of Natural Resources, Illinois Environmental Protection Agency and Illinois Department of Natural Resources, Minnesota Pollution Control Agency, and Wisconsin Department of Natural Resources. We detailed procedures for initial data selection in **Supplementary Information**.

We considered watersheds as independent replicates of metacommunities if the river network flows into one of the following: the ocean, a large lake/reservoir ($\ge$ 10 km² in the areal area) or a large river that may represent lentic habitats ($\ge$ 5000 km² in the watershed area). These watersheds are small enough (`r round(min(c(wsd_subset[[1]]$area, wsd_subset[[2]]$area)))` to `r round(max(c(wsd_subset[[1]]$area, wsd_subset[[2]]$area)))` km^2^) to assume that fishes can disperse therein at an ecological time scale[@giamEnvironmentPredationGovern2016], while posing challenges for stream fishes to traverse across watersheds (the ocean or lentic habitats). Metacommunities with $\ge$ 10 sampling sites were used in our analysis because a small sample size inflates statistical uncertainties in estimating diversity metrics (**Supplementary Information**).

We estimated $\alpha$, $\beta$, and $\gamma$ diversity at each watershed. $\gamma$ diversity is the total species richness at each watershed. Since it is impossible to sample all species present in each watershed, we used asymptotic species richness as a proxy for $\gamma$ diversity. Asymptotic species richness can be interpreted as the estimated true species richness in a given area when estimated using spatial replicates of local community data within a metacommunity[@chaoRarefactionExtrapolationHill2014a; @chaseEcosystemDecayExacerbates2020]. We estimated asymptotic species richness using sampling sites located in the same watershed (i.e., spatial replicates within a metacommunity) via the R package iNEXT[@hsiehINEXTPackageRarefaction2016]. We assessed the completeness of the observed species richness based on sample coverage, which is the proportion of observed species richness to asymptotic species richness. We excluded watersheds whose sample coverage was below 0.9 because the estimates of asymptotic species richness may contain non-negligible statistical uncertainty. $\alpha$ diversity was estimated by taking an average of local species richness at each site. $\beta$ diversity was estimated by dividing the estimated $\gamma$ diversity by $\alpha$ diversity.

@@

We collected data for watershed characteristics (watershed area [km^2^], branching probability, and mean elevation [m]), climatic variables (annual mean air temperature [degree celsius] and cumulative precipitation [mm]), and land use (fraction of forest, fraction of urban, fraction of agriculture, and dam density per watershed area [km^-2^]). We delineated watershed polygons and river lines (defined as $\ge$ 1 km^2^ in the watershed area) using Hydrologically Adjusted Elevations of MERIT Hydro[@yamazakiMERITHydroHighResolution2019] with ArcMap 10.7 and SAGA 6.3.0 combined with the R package 'RSAGA'[@brenningRSAGASAGAGeoprocessing2018]. Watershed area was estimated based on the delineated watershed polygons using functions in the R package 'sf'[@pebesmaSimpleFeaturesStandardized2018]. We estimated branching probability by fitting exponential distributions to histograms of branch length (unit: kilometer, river segment between successive confluences or a confluence and the outlet/upstream terminal) as $branch~length \sim Exp(\lambda)$. A derived statistical quantity $1 − e^{−\lambda}$ corresponds to the theoretical branching probability $P_b$. We obtained mean elevations for each watershed polygon using Hydrologically Adjusted Elevations of MERIT Hydro[@yamazakiMERITHydroHighResolution2019]. We estimated the spatial averages of air temperature and cumulative precipitation at a 1-km resolution for each watershed, which were extracted from WorldClim 1.4[@hijmansVeryHighResolution2005] via the R package 'raster.' We calculated fraction of forest, urban, and agriculture for each watershed using land use data in 2015 from Copernicus Global Land Service (100-m resolution)[@marcelbuchhornCopernicusGlobalLand2020]. We calculated dam density per watershed area (km^-2^) based on the Global Reservoir and Dam Database, Version 1 (GRanDv1)[@lehnerGlobalReservoirDam2011; @lehnerHighResolutionMappingWorld2011].

@@

We developed linear models to examine the effects of macro-scale drivers on diversity metrics. Log-transformed diversity metrics ($\alpha$, $\beta$ and $\gamma$ diversity) at watershed $i$, $log_{10}~y_i$, are assumed to follow a normal distribution as $log_{10}~y_i \sim Normal(\mu_i, \sigma^2)$. The parameter $\mu_i$ was related to linear predictors, including log-transformed watershed area (ecosystem size; $log_{10}~WA$), log-branching probability (ecosystem complexity; $log_{10}~P_b$), region (coded as Hokkaido = 0, Midwest = 1; $region$), air temperature, precipitation, and fraction of forest land use (logit-transformed). Although our primary focus of this analysis is to assess the effects of watershed area and branching probability, we included climate and land use variables to control these effects statistically. Fractions of urban and agriculture land use were not included in the statistical models because these variables are strongly correlated with the fraction of forest land use (**Figure S19**). Also, we did not use mean elevation in the statistical models because it showed a moderate correlation with air temperature (**Figure S19**). Correlations for other combinations of explanatory variables were weak to moderate.

Climate (air temperature and precipitation) and land use (fraction of forest and dam density) differed clearly between the regions. Therefore, we used residuals of the simple linear regressions between the climate or land use variable (response variable) and region (explanatory variable). The residuals represent deviations from the regional averages; positive values indicate values greater than the regional average while negative values indicating values smaller than the regional average. Therefore,

@@

$$
\mu_i = \xi_0 + \xi_1~log_{10}~WA_i + \xi_2~log_{10}~P_{b,i} + \xi_3~region_i + \sum_k \xi_k x_{k,i}
$$
where $\xi_k$ are the intercept and regression coefficients for linear predictors. Meanwhile, the region-specific model assumes the effects of watershed area and branching probability differ between regions (i.e., interactions between region and watershed area/branching probability).

@@

# Acknowledgements
We acknowledge numerous scientists who have been involved in the collection of fish community data. We thank Michio Fukushima for providing information on the Hokkaido Freshwater Fish Database HFish.

@@

# Data and code availability
Data are provided by third parties and references are provided in the main text where appropriate. R functions for the generation of branching networks and the metacommunity simulation are provided as the R package 'mcbrnet.' (available at https://github.com/aterui/mcbrnet). Codes for simulations, statistical analysis, figures, and tables are available at https://github.com/aterui/public-proj_stream-diversity.
